---
title: "Explore and Summarize Data"
author: "Andrew Marksberry"
date: "4/27/2021"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#Load all packages neccessary to complete the analysis

knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
library(ggplot2)
library(grid)
library(gridExtra)
library(GGally)
library(dplyr)
library(tidyr)
```

```{r echo=FALSE, Load_the_Data}
#Load the Data
white_Wine_data <- read.csv('wineQualityWhites.csv',row.names=NULL)
white_Wine_data$X <- NULL
```

# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots}
str(white_Wine_data)

white_Wine_data$quality.level <- ifelse(white_Wine_data$quality > 7, "high", 
                             ifelse(white_Wine_data$quality > 4, "average", "low"))
white_Wine_data$quality.level <- factor(white_Wine_data$quality.level, 
                             levels=c("high", "average", "low"), ordered=TRUE)
attach(white_Wine_data)
```

The dataset consist of 12 variables and 4898 observations.

```{r echo=FALSE}
qplot(factor(quality), data=white_Wine_data, geom="bar", xlab="Quality") + theme_bw()
summary(white_Wine_data$quality)
```

White wine quality is normally distributed and concentrated around 5 and 6.

```{r echo=FALSE}
uni_qplot <- function(x, dat=NULL, xlab, binwidth=0.01) {
  if(missing(dat)) {
    qplot(x, data=white_Wine_data, xlab=xlab, binwidth=binwidth) + theme_bw()
  }
  else {
    qplot(x, data=dat, xlab=xlab, binwidth=binwidth) + theme_bw()
  }
}
uni_qplot(x=fixed.acidity, xlab="Fixed acidity (g/dm^3)", binwidth=0.1)
summary(fixed.acidity)
```

The distribution of fixed acidity is right skewed, and concentrated around 6.8 

```{r echo=FALSE}
uni_qplot(x=volatile.acidity, xlab="Volatile acidity (g/dm^3)")
summary(white_Wine_data$volatile.acidity)
```

The distribution of volatile acidity is also right skewed and concentrated around .26.

```{r echo=FALSE}
uni_qplot(citric.acid, xlab="Citric acid (g/dm^3)")
summary(white_Wine_data$citric.acid)
```

The distribution of citric acid is normal and concentrated around .32 and the 
data contains a few outliers.

```{r echo=FALSE}
uni_qplot(residual.sugar, xlab="Residual sugar (g/dm^3)", binwidth=0.5)
summary(white_Wine_data$residual.sugar)
```

The distribution of residual sugar is right skewed.

```{r echo=FALSE}
uni_qplot(chlorides, xlab="Chlorides (g/dm^3)")
summary(white_Wine_data$chlorides)
```

The distribution of chlorides is normal and is concentrated around .4.

```{r echo=FALSE}
qplot(free.sulfur.dioxide, xlab="Free sulfur dioxide (g/dm^3)")
summary(white_Wine_data$free.sulfur.dioxide)
```

The distribution of free sulfur dioxide is normal.

```{r echo=FALSE}
uni_qplot(total.sulfur.dioxide, xlab="Total sulfur dioxide (g/dm^3)", binwidth=.5)
summary(white_Wine_data$total.sulfur.dioxide)
```

The distribution of total sulfur dioxide seems to be unclear whether it is bimodel 
or unimodel, right skewed or normal. 

```{r echo=FALSE}
uni_qplot(density, xlab="Density (g/dm^3)")
summary(white_Wine_data$density)
```

The distribution of density is right skewed and concentrated at .5

```{r echo=FALSE}
uni_qplot(pH, xlab="pH (g/dm^3)")
summary(white_Wine_data$pH)
```

The distribution of pH is normal and is concentrated around 3.1

```{r echo=FALSE}
uni_qplot(sulphates, xlab="Sulphates (g/dm^3)")
summary(white_Wine_data$sulphates)
```

The distribution of sulphates is normal and concentrated around .5

```{r echo=FALSE}
uni_qplot(alcohol, xlab="Alcohol (g/dm^3)", binwidth=.5)
summary(white_Wine_data$alcohol)
```

The distribution of alcohol is right skewed and is concentrated around 9.5

Next I divided the data into two groups byt quality.  High quality group contains 
the observations receiving 7 or 8, and the low quality group which contains the observations receiving 3 or 4.
After looking at the data there may be a correlation with quality based in the volatile acid, density, and citric acid. 
The graphs below will show the data differences.

```{r echo=FALSE}
highquality <- subset(white_Wine_data, quality== 8 | quality == 7)
lowquality <- subset(white_Wine_data, quality == 3 | quality == 4)
highvolatile <- uni_qplot(highquality$volatile.acidity, dat=highquality,
                          xlab="Volatile acidity, High Quality",
                          binwidth=0.1)
lowvolatile <- uni_qplot(lowquality$volatile.acidity, dat=lowquality,
                         xlab="Volatile acidity, Low Quality",
                         binwidth=0.1) 

highdensity <- uni_qplot(highquality$density, dat=highquality,
                         xlab="Density, High Quality", binwidth=0.001)
lowdensity <- uni_qplot(lowquality$density, dat=lowquality,
                        xlab="Density, Low Quality", binwidth=0.001)

highcitric <- uni_qplot(highquality$citric.acid, dat=highquality,
                        xlab="Citric Acid, High Quality")
lowcitric <- uni_qplot(lowquality$citric.acid, dat=lowquality,
                       xlab="Citric Acid, Low Quality")

highalcohol <- uni_qplot(highquality$alcohol, dat=highquality, 
                       xlab="Alcohol, High Quality", binwidth=0.1)

lowalcohol <- uni_qplot(lowquality$alcohol, dat=lowquality, 
                         xlab="Alcohol, Low Quality", binwidth=0.1)

grid.arrange(lowvolatile, highvolatile, lowdensity, highdensity,
             lowcitric, highcitric, lowalcohol, highalcohol, ncol=2)
```

# Univariate Analysis

### What is the structure of your dataset?

There are 4,899 white wines that where tested with 12 features on the property 
of the wine. (fixed.acidity, volatile.acidity, citric.acid, residual.sugar, chlorides,
              free.sulfur.dioxide, total.sulfur.dioxide, density, pH, sulphates, alcohol, and quality)

Other observations:
The median quality of white wine rating is 6.  Most of the wines have a density around .5. 
The median percent of alcohol is 10.4 and the max is 14.2. 

### What is/are the main feature(s) of interest in your dataset?

My main feature of interest is the wine quality and if there is a connection
between the quality and some or all of the collected items.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

I beleive that alcohol, acidity, and residual.sugar will have a relationship to perceived quality.

### Did you create any new variables from existing variables in the dataset?

I created a variable called quality.level that will divide the quality into categories 
or low, average, and high.  This will make dividing up the data by quality easier.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

The citric acid contained some unusual distributions.  The data was already clean so no modifications or cleaning was necessary.

# Bivariate Plots Section

```{r Bivariate_Plots}
bi_qplot <- function(x, y, z="jitter") {
  if(z=="boxplot") {
    qplot(x=x, y=y, data=white_Wine_data, geom="jitter", alpha=0.01) + 
      geom_boxplot() +
      guides(alpha="none") +
      theme_bw()
  }
  else {
    qplot(x=x, y=y, data=white_Wine_data, geom=z, alpha=0.01) + 
      guides(alpha="none") +
      theme_bw()
  }
}

bi_qplot(quality.level, volatile.acidity, "boxplot") +
  xlab("Quality level") +
  ylab("Volatile acidity (g/dm^3)")
```

The graph shows that the high and average quality wines have lower volatile acidty than the lower rated wines.

```{r}
bi_qplot(quality.level, alcohol) +
  xlab("Quality level") +
  ylab("Alcohol")
cor.test(white_Wine_data$quality, white_Wine_data$alcohol)
```

With the correlation coefficient of 0.435, the graph shows a positive relationship 
between alcohol and quality level. Average quality and low quality wines have their 
percent alcohol contents concentrated around 10 whereas high quality wines have their 
percent alcohol contents concentrated around 12.

```{r}
bi_qplot(citric.acid, volatile.acidity) +
  xlab("Citric acid (g/dm^3)") +
  ylab("Volatile acidity (g/dm^3)")
cor.test(white_Wine_data$citric.acid, white_Wine_data$volatile.acidity)
```

There is a negative correlation between citric acid and volatile acidity.

```{r}
bi_qplot(alcohol, density) + 
  xlab("Alcohol (%)") + 
  ylab("Density (g/cm^3)")
cor.test(white_Wine_data$alcohol, white_Wine_data$density)
```
The correlation coefficient is -0.7, this shows that as the alcohol content increases the density decreases.
# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

I observed a negative relationship between citric acid and volatile acidity, and 
a positive coorelation between quality level and alcohol.

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

Yes, I observed that as alcohol content increases the density decreases.

### What was the strongest relationship you found?

Quality is positively correlated with the alcohol content.

# Multivariate Plots Section

```{r Multivariate_Plots}
multi_qplot <- function(x, y, z, alpha=0.4) {
  qplot(x, y, data=white_Wine_data, color=z) +
    geom_point(position=position_jitter(w = 0.025, h = 0.025), alpha=alpha) +
    guides(alpha=FALSE)
}
multi_qplot(density, volatile.acidity, quality.level) +
  xlab("Density (g/cm^3)") +
  ylab("Volatile acidity (g/cm^3)") +
  labs(color="Quality level")
```

The densities of the high quality wines are concentrated between .99 and 1.

```{r}
multi_qplot(volatile.acidity, alcohol, quality.level) + aes(size=citric.acid) +
  xlab("Volatile acidity (g/dm^3)") +
  ylab("Alcohol (%)") + 
  labs(color="Quality level", size="Citric acid")
print("Percent alcohol contents by quality level:")
white_Wine_data %>% 
  group_by(quality.level) %>% 
  summarize(mean=mean(alcohol),sd=sd(alcohol))
print("Volatile acidities by quality level:")
white_Wine_data %>% 
  group_by(quality.level) %>% 
  summarize(mean=mean(volatile.acidity),sd=sd(volatile.acidity))
```

High quality feature seems to be associated with alcohol ranging from 12 to 14, 
volatile acidity from 0.3 to 0.6, and citric acid from 1.2 to 1.6.

```{r}
multi_qplot(fixed.acidity, volatile.acidity, quality.level) + 
  aes(size=pH) +
  xlab("Fixed acidity (g/dm^3)") +
  ylab("Volatile acidity (g/dm^3)") +
  labs(color="Quality level")

multi_qplot(residual.sugar, alcohol, quality.level) + 
  xlab("Residual sugar (g/dm^3)") +
  ylab("Alcohol (%)") +
  labs(color="Quality level")

multi_qplot(fixed.acidity, alcohol, quality.level) + 
  aes(size=citric.acid) +
  xlab("Fixed acidity (g/dm^3)") +
  ylab("Alcohol (%)") + 
  labs(color="Quality level", size="Citric acid")

den_qplot <- function(x, color, xlab) {
  ggplot(data=white_Wine_data, aes(x, colour=color)) + 
    geom_density() + 
    xlab(xlab) + 
    labs(colour="Quality level") +
    theme_bw()
}
den_qplot(fixed.acidity, quality.level, "Fixed acidity (g/dm^3)")
```

The distribution of low and average quality wines seem to be concentrated at fixed 
acidity values that are between 4 and 7. Citric acid increases as fixed acidity increases.

```{r}
alcoQuaDensity <- den_qplot(alcohol, quality.level, "Alcohol (%)")
print(alcoQuaDensity)

alcohol_lm <- lm(data=white_Wine_data, quality~alcohol)
summary(alcohol_lm)
```

High quality wine density line is distinct from the others, and mostly distributed between 12 and 13.

```{r}
volaQuaDensity <- den_qplot(volatile.acidity, quality.level, 
                            "Volatile acidity (g/dm^3)")
print(volaQuaDensity)

volacid_lm <- lm(data=white_Wine_data, quality~volatile.acidity)
summary(volacid_lm)
```

This chart shows a clear trend; as volatile acidity decreases, the quality 
of wine increases. Wines with volatile acidity exceeding 0.6 are mostly rated as low 
quality.

# Multivariate Analysis

Quality has a strong positive relationship with the % of alcohol and the density. 
It may be possible to make a predictive model that can predict the quality of wide
based on % of alcohol and density.

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

When looking at wine quality level, there is a positive relationship between density and alcohol.

### Were there any interesting or surprising interactions between features?

I feel like residual sugar should have an effect on the flavor but it had little impact on wine quality.

# Final Plots and Summary

### Plot One

```{r Plot_One}
ggplot(data=white_Wine_data, aes(factor(quality), fill=quality.level)) + 
  geom_bar() + 
  xlab("Quality") + 
  ylab("Number of wines")
```

### Description One

The distribution of white wine quality is normal.  The average quality wines make up 74.6% of the observations.  There are no wines ranked 1,2, or 10.

### Plot Two

```{r Plot_Two}
bi_qplot(quality.level, citric.acid, "boxplot") +
  xlab("Quality level") +
  ylab("Citric Acid (g/dm^3)")
```

### Description Two

Citric Acid seems to have no correlation with quality. It is surprising that the
average for all three quality levels appears to be the same.  The acidity of a 
wine can greatly change the flavor but it did not seem to effect the results.

### Plot Three

```{r Plot_Three}
vol.alco <- multi_qplot(volatile.acidity, alcohol, quality.level) + 
  geom_point(size=4, shape=2, colour="steelblue", alpha=0.002) + 
  xlab("Volatile acidity (g/dm^3)") +
  ylab("Alcohol (%)") +
  labs(color="Quality level") +
  scale_colour_grey() + 
  theme_bw()

# Move to a new page
grid.newpage()
# Create layout : nrow = 2, ncol = 2
pushViewport(viewport(layout = grid.layout(2, 2)))
# A helper function to define a region on the layout
define_region <- function(row, col){
  viewport(layout.pos.row = row, layout.pos.col = col)
} 
# Arrange the plots
print(vol.alco, vp=define_region(1, 1:2))
print(volaQuaDensity, vp = define_region(2, 1))
print(alcoQuaDensity, vp = define_region(2, 2))

```

### Description Three

It appears that white wines that are rated higher have a higher alcohol content 
and lower volatile acidity.  In conclusion alcohol and volatile acidity are an 
important factor in the quality of whie wines.

###Reflection

The white wines data set contains ratings on 4898 white wines.  To understand
the dataset I first graphed the provided data to find questions that I wanted
to answer.  After figuring out the questions I wanted to answer I created graphs to 
provide me with my answers.

There is a trend between the alcohol content and the quality of the wine. There
is also a trend between the volatile acidity and the quality of the wine.

With the observations provided average quality made up almost 75% of the observations.
In order to run further analysis there could be double the observations, or a larger data
set for individuals to answer.  Maybe add age, country, or state to be collected
and see if there is correlation to rankings.

